---
title: "Project 2, parts 4 and 5: RNA-seq analysis"
output: html_document
date: "2025-05-08"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Project 2 parts 4 and 5 overall: 75 points

Include any code that you used to solve the question as well as the
answers for all below questions. In parts 4 and 5 from project 2, you
will explore how gene expression differs between electric fish tissues.
You will learn how to analyze RNA-seq count data: identify
differentially expressed genes, distinguish samples, and make biological
interpretations. Additionally, you will visualize findings and
communicate them effectively. **Note: Please submit the Rmd and the knit
HTML file for this assignment**

### Dataset reminder:

In this project, you are be analyzing the count data and identifying
differentially expressed genes between skeletal muscle and electric
organ tissue. As a reminder, these datasets originated from two
different electric fish studies (PRJNA1005245 and PRJNA1005244). It may
help to consider how these papers analyzed the data: the methods for the
PRJNA1005244 dataset are
[published](https://doi.org/10.1093/molbev/msae021) and the methods for
the PRJNA1005245 dataset are written in the third chapter of a
[thesis](https://canvas.uoregon.edu/courses/266187/files/22059308?module_item_id=5380118).
These studies are both using the genus *Campylomormyrus*. **For this
assignment, you may not need to analyze all of the data to address the
hypothesis.**

### Resources:

There are several resources that you may use to complete this
assignment:

1)  A primary resource: [DESeq2
    workflow](https://master.bioconductor.org/packages/release/workflows/vignettes/rnaseqGene/inst/doc/rnaseqGene.html#pre-filtering-the-dataset)

2)  [DESeq2's additional
    guide](https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html)

3)  [Harvard Bioinformatics Core
    training](https://hbctraining.github.io/DGE_workshop/lessons/04_DGE_DESeq2_analysis.html)

You are expected to use DESeq2 for this assignment. If you decide to use
an additional package, please list the package name, how you found it,
and why you chose to use it. We will do our best to help you navigate
other packages, but we cannot guarantee support for other packages.

## Project 2 part 4: Hypothesis testing and exploration

0.  In part 4, your goal is to investigate differences in gene
    expression between the electric organ and skeletal muscle. Write the
    null hypothesis for this goal below.

```         
Null Hypothesis: There is no difference in gene expression between tissue samples
```

1.  Load your data into DESeq2

- Your data is located on Talapas:
  /projects/bgmp/shared/Bi623/Project2/counts.tar
- **Think carefully about your design formula and explain your chosen
  formula below**
- Note: you can follow the Bear_analysis rmd from class as a template or
  can also reference other resources online, such as this guide from
  [Duke](https://people.duke.edu/~ccc14/duke-hts-2017/Statistics/08032017/DESeq2-Notebook-introduction.html)
  to load your data.

```{r}
#I found these packages on the Duke website
library(tidyverse)
library(edgeR)
library(limma)
library(Glimma)
library(ggplot2)
library(RColorBrewer)
library(DESeq2)
library(apeglm)
```

```{r}
#data load in
campy_meta = read.csv("C:/Users/kylie/Documents/Project2_P4/Project2_RNAseq_metadata_campy.csv")

campy_filtered = filter_out(campy_meta, campy_meta$AgeSize == "6cm")

campy = DESeqDataSetFromHTSeqCount(campy_filtered, directory="C:/Users/kylie/Documents/Project2_P4/counts_tar", design=~ Species+Tissue+Species:Tissue)
```

```         
The experiment is filtered by species since that is also a variable relevant to the design.
```

2.  Filter the dataset to remove genes that are lowly expressed

- There are multiple ways to filter RNA-Seq datasets. You can either use
  cpm(counts per million) as a filter as we did in the Bear_analysis or
  use raw counts completed in the DESeq2 vignette from the above
  resources.
- Describe why you used your filtering strategy

```         
xxx
```

```{r}
mycpm = cpm(campy)
head(mycpm)
```

[Specify which approach you chose and why]

3.  Normalize the dataset and run differential expression analysis

- You may choose to normalize and run differential expression separately
  as done in the Bear_analysis, or in 1 command as done in the DESeq
  workflow 5.1.

```{r}
thresh = mycpm > 0.5

head(thresh)
```

```{r}

table(rowSums(thresh))

# we would like to keep genes that have at least 2 TRUES in each row of thresh
keep <- rowSums(thresh) >= 2
# Subset the rows of count data to keep the more highly expressed genes
campy <- campy[keep,]

## Examining how many genes we have left in the dataset
campy
```

4.  Normalize the data to use for dimensionality reduction

- Use the rlog transformation method (follow along with the workflow,
  section 4.2)
- Describe why the rlog transformation is performed

```         
The rlog is preformed because data needs be normalized for genes with very low counts because those numbers inflate the varience.
```

```{r}
campy <- estimateSizeFactors(campy)

## Retrieving size factors
sizeFactors(campy)

## Comparing size factors to library size
colSums(counts(campy))


campy = DESeq(campy)



```

```{r}
rld <- rlog(campy, blind = FALSE)
head(assay(rld), 3)
```

5.  Identify distances between samples using sample distance heatmaps

- Generate a Euclidean distance heatmap and dendrogram using your rlog
  transformed data (follow workflow section 4.3)
- Create a Poisson distance heatmap with dendrogram using DESeq data
  (follow workflow section 4.3)
- Describe what each of these heatmaps are measuring
- Look at the sample groupings for each heatmap+dendrogram. Do you
  notice any potential individual outliers? Do both plots agree?

Note: You may need to install the packages “pheatmap” and “RColorBrewer”
to follow along. If R tells you to restart your session listen to it.

```{r}
library(pheatmap)
library(RColorBrewer)
```

```{r}
sampleDists <- dist(t(assay(rld)))
sampleDists
```

```{r}
sampleDistMatrix <- as.matrix( sampleDists )
rownames(sampleDistMatrix) <- paste( rld$Tissue, rld$SRRfile, sep = " - " )
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors)
```

```{r}
library("PoiClaClu")
poisd <- PoissonDistance(t(counts(campy)))
```

```{r}
samplePoisDistMatrix <- as.matrix( poisd$dd )
rownames(samplePoisDistMatrix) <- paste( campy$Tissue, campy$SRRfile, sep=" - " )
colnames(samplePoisDistMatrix) <- NULL
pheatmap(samplePoisDistMatrix,
         clustering_distance_rows = poisd$dd,
         clustering_distance_cols = poisd$dd,
         col = colors)
```

[Describe the two heatmaps and what they are measuring (Any outliers? Do
the plots have the same interpretation?)]

```         
There is one outlier in the electric organ samples with one random skeletal muscle sample (ending in 366) being lumped in so that could be an outlier that is left out.
```

6.  Examine similarity in sample global gene expression using PCA,
    Generalized PCA, and MDS plots.

- Generate PCA, Generalized PCA, and MDS plots (following workflow
  section 4.4-4.6)
- Modify the color and shape of points to reflect Tissue and Species of
  samples
- Briefly describe each approach and their findings.

```{r}
pcaData <- plotPCA(rld, intgroup = c("Tissue", "SpeciesSmaller"), returnData = TRUE)
pcaData
```

```{r}
percentVar <- round(100 * attr(pcaData, "percentVar"))
```

```{r}
ggplot(pcaData, aes(x = PC1, y = PC2, color = Tissue, shape = SpeciesSmaller)) +
  geom_point(size =3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed() +
  ggtitle("PCA with RLD data")
```

```{r}

library("glmpca")
gpca <- glmpca(counts(campy), L=2)
gpca.dat <- gpca$factors
gpca.dat$Tissue <- campy$Tissue
gpca.dat$SpeciesSmaller <- campy$SpeciesSmaller

ggplot(gpca.dat, aes(x = dim1, y = dim2, color = Tissue, shape = SpeciesSmaller)) +
  geom_point(size =3) + coord_fixed() + ggtitle("glmpca - Generalized PCA")
```

```{r}
mds <- as.data.frame(colData(rld))  %>%
  cbind(cmdscale(sampleDistMatrix))
ggplot(mds, aes(x = `1`, y = `2`, color = Tissue, shape = SpeciesSmaller)) +
  geom_point(size = 3) + coord_fixed() + ggtitle("MDS with RLD data")
```

7.  Use your dimensionality reduction plots, evaluate sample quality and
    complete quality adjustments.

- Based on your dimensionality reduction plots, do you have concerns
  about any samples? Explain.
- If so, remove them now, repeat steps of loading in data, filtering,
  normalization, and running the differential expression analysis.

[Describe your answer. If you identify a problem, then include the code
used to repeat the initial data loading, filtering, normalization, and
DE steps]

```{r}
campy_meta2 = filter_out(campy_filtered, SRRfile == "SRR25630366")

final_campy = DESeqDataSetFromHTSeqCount(campy_filtered, directory="C:/Users/kylie/Documents/Project2_P4/counts_tar", design=~ Species+Tissue+Species:Tissue)
myCPM2 = cpm(final_campy)

thresh2 = myCPM2 > 0.5

# Summary of how many TRUEs there are in each row by sample
table(rowSums(thresh2))

# we would like to keep genes that have at least 2 TRUES in each row of thresh
# I.e. we keep any gene with myCPM values above .5 for 2 or more samples
keep2 <- rowSums(thresh2) >= 2
# Subset the rows of count data to keep the more highly expressed genes
final_campy = final_campy[keep2,]
```
